---
title: "Intro to Data Frames in R"
subtitle: "Environmental Data Analytics"
author: "Luana Lima"
date: "Fall 2023"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data frame definition

A **data frame** is used for storing data tables. It a list a vector of equal length.

When we import data to R, data frame is the preferred way for storing the data because columns can have different modes (character, numeric, integer, logical, complex).

## Data frame built-in example

Let's look into a built-in data frame from package "datasets" - cars. The data give the speed of cars and the distances taken to stop.

```{r cars}
cars
#by issuing the name of the dataframe you can preview it
mtcars
#the output is within the document, chunk output inline, or in colse, chunk output in console, when issued in console the command opens a new tab to preview the table
```


Note that it has 2 columns and 50 rows.

## Data frame columns

Suppose you want just the column speed. How would you access that data?

```{r}
cars$speed #returns a vector of the values in the speed column 
mtcars$cyl
```

How would you store it on another object?

```{r}
car_speed <- cars$speed
car_speed
```

## Transforming object in a data frame

Is the new object you create a data frame?

```{r}
#Option 1
class(car_speed)

#Option 2
is.data.frame(car_speed)
is.vector(car_speed)
```

How could you make it a data frame?

```{r}
df_car_speed <- as.data.frame(car_speed) #creates a dataframe from vectors
df_car_speed
class(df_car_speed)
```

## Adding columns to a data frame

How could you add columns to *df_car_speed*?

```{r}
car_dist <- cars$dist

#Option 1
df <- cbind(df_car_speed,car_dist)  #similarly rows could be added using rbind(), binds columns and makes a dataframe if they are the same length
class(df)
df

#Or Option 2 - transform into a data frame before binding
df_car_dist <- as.data.frame(car_dist)  #op2
df_opt2 <- cbind(df_car_speed,df_car_dist) #you can also combine two dataframes (instead of two vectors)
class(df_opt2)
df_opt2

```

Note that when we transformed the vector in a data frame the name of the vector became the column name.

```{r}
colnames(df)  #or simply names()
names(df)
```

## Creating a data frame

How would you create a data frame?

```{r}
#useful function data.frame()
create_df <- data.frame("speed of car"=car_speed,"dist of car"=car_dist) #creating new dataframe rather than transforming something into a dataframe
create_df
```

## Data frame functions

Some useful functions to use with data frames.

```{r}
ncol(df)  
nrow(df)  
length(df)  #same as ncol
summary(df)
head(df,2) #show the first 2 rows of df

#If you know the number of the column you want you can refer to that to access column
df[,1] #all rows, only the first column
df[1,1] #first row and column
df[1,] #all columns for the first row
#slicing
df[1:10,2] #column 2, rows 1-10
df2<- df[c(2,1,3),] #specifying rows two, one, and three, and all of the columns will display (this notation only applies to a dataframe)
df2$car_speed #returns the column car speed for our new dataframe
class(df2)


#you could also use this notation to delete columns
df <- df[,-2]
df
```
