---
title: "Assignment 5: Data Visualization"
author: "Emma Kaufman"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
#Install packages
library(tidyverse)
library(lubridate)
library(here) #The here package allows for better control of relative paths
library(cowplot)

#Ensure that "here" points to your project folder
here()

#get working directory
getwd()

#read in files
LakeChem <- read.csv(
  file=here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), 
  stringsAsFactors = TRUE
)

Litter <- read.csv(
  file=here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), 
  stringsAsFactors = TRUE
)

#2 
#changing to date format
Litter$collectDate <- ymd(Litter$collectDate)
LakeChem$sampledate <- ymd(LakeChem$sampledate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
my_theme <- theme_base()  + 
  theme(
    line = element_line(
      color = 'navyblue',
      linewidth = 2
      ),
    plot.title =        element_text(
      color = 'navyblue',
      size = 15
    ),
    axis.text =         element_text(
      color = 'navyblue'
    ),
    axis.title.x = element_text(
      color = "navyblue"),
  axis.title.y = element_text(
    color = "navyblue"),
    plot.background = element_rect(
      fill = 'lightblue'
    ))

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
phos_plot<- 
  ggplot((LakeChem),
         aes(
           x = tp_ug,
           y = po4,
           shape = lakename
         )) +
  geom_point()+
  geom_smooth(method = 'lm', 
              color = 'black')+
  ylim(0,50)+
  ggtitle("Total Phosphorus by Phosphate for Peter and Paul Lakes")+
  my_theme

print(phos_plot)

#do i need two trendlines?

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 

unique(LakeChem$month) #seeing which months should plot data 

#a) plotting temperature
Temperatureplot <-
  ggplot((LakeChem), 
         aes(
           x = factor(LakeChem$month, levels=1:12, labels = month.abb), 
           y = temperature_C, 
           color = lakename)) +
  geom_boxplot() +
  ggtitle("Temperature for Peter and Paul Lakes") +
  scale_x_discrete(name= "month", drop = FALSE) +
  labs(y = "temperature (C)") +
    my_theme +
  theme(legend.position = "bottom")

  
print(Temperatureplot)

#b) plotting total phosphorus
TPplot <-
  ggplot((LakeChem), 
         aes(
           x = factor(month, levels=1:12, labels = month.abb), 
           y = tp_ug, 
           color = lakename)) +
           #color = factor(LakeChem$lakename))) +
  geom_boxplot() +
  ggtitle("Total Phosphorus for Peter and Paul Lakes") +
  scale_x_discrete(name= "month", drop = FALSE) +
  labs(y = "phosphorus (ug)")+
  my_theme 

print(TPplot)

#c) plotting total nitrogen
TNplot <-
  ggplot((LakeChem), 
         aes(
           x = factor(month, levels=1:12, labels = month.abb), 
           y = tn_ug, 
           color = lakename)) +
  geom_boxplot() +
  ggtitle("Total Nitrogen for Peter and Paul Lakes")+
  scale_x_discrete(name= "month", drop = FALSE) +
  labs(y = "nitrogen (ug)") +
  my_theme 

print(TNplot)

plot_grid(TNplot+theme(legend.position = "none"),
          TPplot+ theme(legend.position = "none"),
          Temperatureplot, ncol = 1,
          rel_heights = c(1,1,1.25),
          align = "vh")
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperature is similar in peter and paul lakes, phosphorus and nitrogen are higher in peter lake. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
Needles_mass <-
  Litter %>% 
  filter(functionalGroup == "Needles")
  

Needlesplot <- ggplot((Needles_mass),
                      aes(
                        #x= factor(collectDate), #what is the form the date should take?
                        x = collectDate,
                        y= dryMass,
                        color = nlcdClass
                      )) +
  geom_point()+ #what is the type of plot I should make?
  ggtitle("Needles dry mass over time")+
  scale_x_date(date_breaks = "4 months", date_labels = '%m/%y')+
  my_theme
print(Needlesplot)

#7
Needlesplot_faceted <- ggplot((Needles_mass),
                      aes(
                        x= collectDate, #what is the form the date should take?
                        y= dryMass
                      )) +
  geom_point()+ #what is the type of plot I should make?
  facet_wrap(vars(nlcdClass), nrow = 3) +
  ggtitle("Needles dry mass over time")+
  scale_x_date(date_breaks = "4 months", date_labels = '%m/%y')+
  my_theme

print(Needlesplot_faceted)

#what do i need to do to dates?

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:the faceted plot because you can see themes